---
title: "EU and EEA Covid-19 Deaths Analysis"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE,
                      message = FALSE)

library(ggplot2)
library(gridExtra)
library(lubridate)
library(tidyverse)
library(dplyr)
library(Hmisc)

data <- read.csv("https://opendata.ecdc.europa.eu/covid19/nationalcasedeath_eueea_daily_ei/csv",
                 na.strings = "", fileEncoding = "UTF-8-BOM")

glimpse(data)

data <- data %>%
  select(-c("continentExp")) %>%
  mutate(dateRep = dmy(dateRep),
         countriesAndTerritories = as.factor(countriesAndTerritories),
         geoId = as.factor(geoId),
         countryterritoryCode = as.factor(countryterritoryCode))

```

A data dictionary for the dataset is available [here](https://www.ecdc.europa.eu/sites/default/files/documents/Description-and-disclaimer_daily_reporting.pdf).

#### Definitions:

* "Incidence rate" is equal to new daily cases per 100K individuals. Country population estimates can be found in 'popData2020.' 

* "Fatality rate" is equal to new daily deaths per 100K individuals. Country population estimates can be found in 'popData2020.' 

---

#### 1. Descriptive Statistics

```{r descriptive_stats, fig.width = 8, fig.height = 8}
#adds 'incidence rate' vector to dataframe
data$incidence_rate<-data$cases/(data$popData2020/100000)

#adds 'fatality rate' vector to dataframe
data$fatality_rate<-data$deaths/(data$popData2020/100000)

head(data)

SubSet.1<-subset(data,data$countriesAndTerritories=="Austria"|data$countriesAndTerritories=="Estonia"|
data$countriesAndTerritories=="Germany"|data$countriesAndTerritories=="Greece"|
data$countriesAndTerritories=="Latvia")

SubSet.1<-SubSet.1[order(SubSet.1$dateRep),]

ggplot(SubSet.1,aes(x=as.Date(dateRep),
                  y=incidence_rate,
                  col=(countriesAndTerritories)))+geom_point() + ylim(0,700) + ggtitle("Incidence Rate by Date") + xlab("Date") + ylab("Incidence Rate")

ggplot(SubSet.1,aes(x=as.Date(dateRep),
                  y=fatality_rate,
                  col=(countriesAndTerritories)))+geom_point() + ylim(0,5) + ggtitle("Fatality Rate by Date") + xlab("Date") + ylab("Fatality Rate")

Austria<-subset(data,data$countriesAndTerritories=='Austria')
Total_Austria_Deaths<-sum(Austria$deaths)
Total_Austria_Cases<-sum(Austria$cases)
DeathPerCase_Austria<-Total_Austria_Deaths/Total_Austria_Cases

Estonia<-subset(data,data$countriesAndTerritories=='Estonia')
Total_Estonia_Deaths<-sum(Estonia$deaths,na.rm=TRUE)
Total_Estonia_Cases<-sum(Estonia$cases,na.rm=TRUE)
DeathPerCase_Estonia<-Total_Estonia_Deaths/Total_Estonia_Cases

Germany<-subset(data,data$countriesAndTerritories=='Germany')
Total_Germany_Deaths<-sum(Austria$deaths,na.rm=TRUE)
Total_Germany_Cases<-sum(Germany$cases,na.rm=TRUE)
DeathPerCase_Germany<-Total_Germany_Deaths/Total_Germany_Cases

Greece<-subset(data,data$countriesAndTerritories=='Greece')
Total_Greece_Deaths<-sum(Greece$deaths,na.rm=TRUE)
Total_Greece_Cases<-sum(Greece$cases,na.rm=TRUE)
DeathPerCase_Greece<-Total_Austria_Deaths/Total_Greece_Cases

Latvia<-subset(data,data$countriesAndTerritories=='Latvia')
Total_Latvia_Deaths<-sum(Latvia$deaths,na.rm=TRUE)
Total_Latvia_Cases<-sum(Latvia$cases,na.rm=TRUE)
DeathPerCase_Latvia<-Total_Latvia_Deaths/Total_Latvia_Cases

Country<-c('Austria','Estonia','Germany','Greece','Latvia')
DeathPerCase<-c(DeathPerCase_Austria,DeathPerCase_Estonia,DeathPerCase_Germany,DeathPerCase_Greece,
DeathPerCase_Latvia)

Death_Table<-data.frame(cbind(Country,DeathPerCase))

ggplot(Death_Table,aes(x=Country,
                  y=DeathPerCase, 
                  col=(Country)))+geom_col()+ggtitle('Death Table')

```

#### 2. Inferential Statistics

```{r inferential_stats, fig.width = 9, fig.height = 8}
SubSet.2<-subset(data,data$countriesAndTerritories=="Austria")
SubSet.2<-SubSet.2[order(SubSet.2$dateRep),]
SubSet.3<-subset(data,data$countriesAndTerritories=="Germany")
SubSet.3<-SubSet.3[order(SubSet.3$dateRep),]

hist(SubSet.2$incidence_rate, main="Incidence Rate Frequency, Austria", xlab="Incidence Rate")
summary(SubSet.2$incidence_rate)
hist(SubSet.3$incidence_rate, main="Incidence Rate Frequency, Germany", xlab="Incidence Rate")
summary(SubSet.3$incidence_rate)

prop.test(x=c(mean(SubSet.2$incidence_rate),mean(SubSet.3$incidence_rate)),n=c(100000,100000))


#null hypo is the mean() incidence rate of Austria - the mean incidence rate of Germany = 0 
#justification: 1) it's categorical data (proportions), 2) working with two samples; 3)comparing the data, Germany vs Austria; hence, we use the two-proportion sample test
#Assumptions of the Two Sample Z Proportion Hypothesis Tests 
#The data are simple random values from both the populations
#Both populations follow a binomial distribution
#Samples are independent of each other
#Test results are accurate when np and n(1-p) are greater than 5
#alpha is .05
#fail to reject null hypothesis because p-value of .06591 is greater than the significance level alpha of .05
#therefore, no significant difference between the two countries Austria and Germany
```

#### 3. Correlation
  
```{r correlation, fig.width = 8, fig.height = 8}
plot(x=data$incidence_rate,y=data$fatality_rate,xlim=c(0,270),ylim=c(0,10), main="Fatality Rate by Incidence Rate",xlab="Incidence Rate", ylab="Fatality Rate")

ggplot(data, aes(x=incidence_rate)) +
  geom_boxplot()+xlim(0,50) + ggtitle("Incidence Rate Boxplot") + xlab("Incidence Rate")

ggplot(data, aes(x=fatality_rate)) +
  geom_boxplot()+xlim(0,2) + ggtitle("Fatality Rate Boxplot") + xlab("Fatality Rate")

ggplot(data, aes(x=incidence_rate)) +
  geom_histogram()+xlim(0,300) + ylim(0,300) + ggtitle("Incidence Rate Histogram") + xlab("Incidence Rate")

ggplot(data, aes(x=fatality_rate)) +
  geom_histogram()+xlim(0,10) + ylim(0,300) + ggtitle("Fatality Rate Histogram") + xlab("Fatality Rate")

cor.test(as.numeric(data$fatality_rate),as.numeric(data$incidence_rate), method = 'spearman')

#Chose the Spearman method because of the skewed data and extreme values (outliers)
#The scatterplot, and correlation coefficient of .1, suggest low correlation between fatality rate and incidence rate
#Scaled x limits on plots because, by definition, fatality rates are a fraction of incidence rates
#y limits are the same on the histograms as to compare the shapes of the distributions
```

#### 4. Regression

```{r regression_a, fig.width = 8, fig.height = 8}

# The code below creates a new data frame, 'model_df,' that includes the area,
# GDP per capita, population and population density for the twenty (20)
# countries of interest

twenty_countries <- c("Austria", "Belgium", "Bulgaria", "Cyprus", "Denmark",
                      "Finland", "France", "Germany", "Hungary", "Ireland",
                      "Latvia", "Lithuania", "Malta", "Norway", "Poland",
                      "Portugal", "Romania", "Slovakia", "Spain", "Sweden")

sq_km <- c(83858, 30510, 110994, 9251, 44493, 338145, 551695, 357386, 93030,
           70273, 64589, 65300, 316, 385178, 312685, 88416, 238397, 49036,
           498511, 450295)

gdp_pps <- c(128, 118, 51, 91, 129, 111, 104, 123, 71, 190, 69, 81, 100, 142,
             71, 78, 65, 71, 91, 120)

model_df <- data %>%
  select(c(countriesAndTerritories, popData2020)) %>%
  filter(countriesAndTerritories %in% twenty_countries) %>%
  distinct(countriesAndTerritories, .keep_all = TRUE) %>%
  add_column(sq_km, gdp_pps) %>%
  mutate(pop_dens = popData2020 / sq_km) %>%
  rename(country = countriesAndTerritories, pop = popData2020)

```


```{r regression_b}

total_cases <- data %>%
  select(c(countriesAndTerritories, cases)) %>%
  group_by(countriesAndTerritories) %>%
  dplyr::summarize(total_cases = sum(cases, na.rm = TRUE)) %>%
  filter(countriesAndTerritories %in% twenty_countries) %>%
  select(total_cases)

model_df <- model_df %>%
  add_column(total_cases)

```


```{r regression_c}
#model<-lm(model_df$total_cases ~ model_df$pop + model_df$pop_dens + model_df$gdp_pps)
#summary(model)

model<-lm(total_cases ~ pop + pop_dens + gdp_pps, data=model_df)
model

summary(model)
summary(model)$coefficient

#Dependent Variable vs Individual Independent Variable
model1<-lm(model_df$total_cases ~ model_df$pop, data=model_df)
model1

plot(model_df$total_cases,model_df$pop,main="Population and Total Cases", xlab="Total Cases", ylab="Population")
summary(model1)

model2<-lm(model_df$total_cases ~ model_df$pop_dens)

plot(model_df$total_cases,model_df$pop_dens,main="Population Density and Total Cases", xlab="Total Cases", ylab="Population Density")
summary(model2)

model3<-lm(model_df$total_cases ~model_df$gdp_pps)

plot(model_df$total_cases,model_df$gdp_pps,main="GDP and Total Cases", xlab="Total Cases", ylab="GDP")
summary(model3)

#Population is statistically significant: t value is high, Pr is low and less than the alpha of .05, indicating statistical significance
#r^2 is 0.8994
#A reduced model without population density and GDP might be in order given their low t values and high Pr
#"Model" here appears to be the most appropriate model, given the apparent strong correlation between population total cases
```

  
```{r regression_d}

# The code below defines our 'newdata' data frame for applying our model to the
# population, population density and GDP per capita for two (2). 

newdata <- data.frame(country = c("Luxembourg", "Netherlands"),
                      pop = c(626108, 17407585),
                      gdp_pps = c(261, 130),
                      pop_dens = c(626108, 17407585) / c(2586, 41540))

Lux<-subset(data, data$countriesAndTerritories=="Luxembourg")
Neth<-subset(data, data$countriesAndTerritories=="Netherlands")

Total_Cases_Lux<-sum(Lux$cases)
Total_Cases_Neth<-sum(Neth$cases)
Total_Cases_Lux_Neth_Actual<-c(Total_Cases_Lux,Total_Cases_Neth)
c("The total cases for Luxembourg and Netherlands:" ,Total_Cases_Lux_Neth_Actual)

predictOutput<-round(predict(model,newdata),4)
c("The predicted totals for Luxembourg and Netherlands:" ,predictOutput)

prediction<-round(predict(model,newdata),4)

ResidualLux<-round(Total_Cases_Lux_Neth_Actual[1]-prediction[1],4)
ResidualNeth<-round(Total_Cases_Lux_Neth_Actual[2]-prediction[2],4)
c("The residual for Luxembourg:" ,ResidualLux)
c("The residual for Netherlands:" ,ResidualNeth)

#The predicted total for the Netherlands is fairly close (residual of 1359905 in a country of population 17407585 and actual cases of 8377385); the predicted total for Luxembourg is off (residual of -3268118 in a country of population 626108)
```
